Synthesizing Quotient Lenses

### 2018-02-25 papers/QuotientLenses/main.tex
Made syntax highlighting changes

\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
% % For double-blind review submission, w/ CCS and ACM Reference
% \documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
% % For single-blind review submission, w/o CCS and ACM Reference (max
% submission space)
% \documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
% % For single-blind review submission, w/ CCS and ACM Reference
% \documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true} % For
% final camera-ready submission, w/ required CCS and ACM Reference
% \documentclass[acmsmall]{acmart}\settopmatter{}

@@

\lstset{
language=Caml, basicstyle=\upshape\sffamily,
keywordstyle=\upshape\sffamily\color{dkblue}, keepspaces=true,
framexleftmargin=1ex, framexrightmargin=1ex, showstringspaces=false,
commentstyle=\itshape\rmfamily, columns=flexible,
% BCP: I find this distracting:
% stringstyle=\sffamily\color{dkred},
mathescape
}

@@

Oftentimes, the same data is stored in a variety different formats. For
example, \bibtex{} and EndNote both store bibliography data; spreadsheet data
can be represented as a CSV file or a TSV file; web APIs can return data as a
JSON or
XML object. When different applications manipulate the same data in different formats,
synchronizing the applications amounts to synchronizing their respective data
formats. One way to synchronize data formats is to use \emph{lenses}, \emph{i.e.,} programs that represent pairs of functions for translating data from
format A to B (the \emph{get} direction) and back again (the \emph{put}
direction).  These lenses, when defined using domain-specific
languages like Boomerang~\cite{boomerang} or
BiGul~\cite{DBLP:conf/pepm/KoZH16}, typically come with a variety of guarantees,
the \emph{lens laws}, that help ensure put and get functions are effective
inverses of one another.  
%comprise such synchronization functions, and the ``lens
%laws'' provide restrictions on how these synchronization functions interoperate%.
%Bidirectional languages like Boomerang~\cite{boomerang} and
%BiGul~\cite{DBLP:conf/pepm/KoZH16} use a single expression to encode all of the
%lens's synchronization functions at the same time.

@@

\noindent
Indeed, Boomerang's quotient lenses are structured this way~\ref{boomerang}.
Of course, some of these elements are optional.  For instance, it is
not necessary to document internal and external formats.  However, we do
view it as good engineering practice to do so, as such documentation makes
lenses easier to read, understand, use and maintain.  In this case, there is
significant redundancy here:  External and internal formats
typically have some overlap; external formats and canonizers have related
structure; and the structure of bijective lens computations is largely driven
by the structure of the internal formats.  Hence, overall, a more compact
notation would simplify description and legibility of such computations.
In addition,
the lens and canonizer languages themselves can be somewhat fiddly and
difficult to
master.  For instance, to quotient out whitespace inside, and in between,
fields while also ignoring field order, one must nest canonizing combinators
within one another in a somewhat tricky way.

@@

One way to define the function used by squash is simply to write it
from scratch in OCaml.  However, our system will also synthesize
such functions automatically from pairs of regular languages and
examples.
%
\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Conway, J. H.", "J. H. Conway")}
let s_to_c = get l
\end{lstlisting}
%
The first line above synthesizes a \emph{bidirectional} lens between
\codefont{comma_name} and \codefont{space_name} using the listed example
transformation as a guide.  The second line simply
extracts the get direction transformation from the lens,
which is what we need for squash.

@@

\subsection{QRE Lenses}
QREs allow users to specify regular languages modulo equivalence relations,
a first step in writing many data transformations.  However, writing the
quotient lens itself still takes a fair amount of work.  Each such lens
is constructed from simple primitive combinators including:
%
\begin{itemize}
\item \kw{del} $s$: delete the constant string $s$ in the get direction;
produce the constant string $s$ in the put direction.
\item \kw{ins} $s$: produce the string $s$ in the get direction;
delete the string $s$ in the put direction.
\item \kw{id} $r$: copy the text matching $r$ in both directions \dpw{this was not the exact syntax.  fix this.}
\end{itemize}
%
In addition, such primitives may be combined using the regular operators
Kleene star, alternation and concatenation, or functional composition of
lenses.  Figure~\ref{fig:example-lens} illustrates the use of these
combinators to transform \bibtex{} to EndNote (and vice versa).
A fair amount of trickiness in writing these transformations as users
must keep mental track of what they have parsed so far \emph{in both
directions}, carefully inserting and deleting the relevant syntax on both
sides.

@@

let author_lens : (lens in (",\n" . bib_author) $\Leftrightarrow$ ("\n" . end_author)) =
del ",\nauthor = \""
. ins "\n%A "
. valid_name
. (del " and "
. ins "\n"
. (ins "%A " . valid_name . del " and " . ins "\n")$^*$ . ins "%A "
. valid_name
|| "")

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.7\textwidth]{canonizers-outside}
\caption{Quotient Lens with ``Canonizers at the Edges''\sam{This is a
screenshot from Nate's thesis. Are we allowed to use it?}\bcp{We can
(if we put ``from [Nate's thesis]'' in the caption), but I don't think
we want the full complexity of asymmetric lenses here.  We should draw
the figure again.  (I have the OmniGraffle file for it, so this is
easy if I know how it should look.)}\bcp{Done---what do you think?}
\sam{Yes this is more accurate, except that we don't need a general choose
function. With QREs the kernel language is always contained in the whole
language so we just use the inclusion as the choose function.}\bcp{I can
fix.}} 
\label{fig:attheedges}
\end{figure}

@@

One combinator that we haven't yet introduced is sequential composition.
The $c \semicolon c'$ QRE successively applies
two equivalence relations to a regular language. For example, suppose that
we wish to express a name as

\noindent $\kw{last\_name \cdot ``,\text{''} \cdot wsp^+ \cdot first\_name}$,
or as $\kw{first\_name\cdot wsp^+ \cdot last\_name}$, with

\noindent $\kw{first\_name\cdot wsp^+ \cdot last\_name}$ as the canonical
representation. We also wish to project away the extra whitespace in between
\kw{first\_name} and \kw{last\_name} once this identification has been made.
Then we can express this equivalence relation by a composition of a
\textit{squash} QRE and a QRE that projects away the extra whitespace in between
\kw{first\_name} and \kw{last\_name}:

@@

\noindent This rule says that the concatenation $c \cdot c'$ of QREs $c$ and
$c'$ is well formed only if the concatentions of $W(c)$ and $W(c')$, and
$K(c)$ and $K(c')$ are ambiguous.

@@

\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonizer(R) &=& id_{\mathcal{L}(R)} \\
\canonizer(\project{R}{s})(w) &=& s\\
\canonizer(\squash{c}{c'}{f})(w) &=&
\begin{cases}
\canonizer(c')(f(w)) & \text{if } w \in \mathcal{L}(W(c))\\
\canonizer(c')(w) & \text{otherwise}
\end{cases}\\
\canonizer(\normalize{R}{R'}{f}) &=& f\\
\canonizer(c' \; ; \; c) &=& \canonizer(c') \circ \canonizer(c)\\
\canonizer(c' \cdot c) &=& \canonizer(c') \cdot \canonizer(c)\\
\canonizer(c_1 \sep c_2)(w) &=&
\begin{cases}
\canonizer(c_1)(w) & \text{if } w \in \mathcal{L}(W(c_1))\\
\canonizer(c_2)(w) & \text{if } w \in \mathcal{L}(W(c_2))\\
\end{cases}\\
\canonizer(c^*) &=& \canonizer(c)^* \\
\end{array}
\]
\end{center}
$\canonizer(\perm{c_1, \ldots, c_n}{c})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \newline
= (\canonizer(c_1)(r_1)) \cdot (\canonizer(c)(s_1)) \cdot \ldots \cdot
(\canonizer(c_n)(r_n))$
\caption{QRE canonizers}
\label{fig:canonizers}
\end{figure}

@@

\section{Conclusion and Future Work}
\label{concl}

>>>>

\documentclass[acmsmall,review,anonymous]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}
% % For double-blind review submission, w/ CCS and ACM Reference
% \documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
% % For single-blind review submission, w/o CCS and ACM Reference (max
% submission space)
% \documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
% % For single-blind review submission, w/ CCS and ACM Reference
% \documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true} % For
% final camera-ready submission, w/ required CCS and ACM Reference
% \documentclass[acmsmall]{acmart}\settopmatter{}

@@

\lstset{
language=Caml, 
basicstyle=\upshape\sffamily,
keywordstyle=\upshape\sffamily\color{dkblue},
keepspaces=true,
framexleftmargin=1ex,
framexrightmargin=1ex,
showstringspaces=false,
commentstyle=\itshape\rmfamily,
emph={synth,project,perm,squash,normalize,using,ins,del,lens},
emphstyle=\upshape\sffamily\color{dkblue},
columns=flexible,
% BCP: I find this distracting:
% stringstyle=\sffamily\color{dkred},
mathescape,
 }

@@

Oftentimes, the same data is stored in a variety different formats. For
example, \bibtex{} and EndNote both store bibliography data; spreadsheet data
can be represented as a CSV file or a TSV file; web APIs can return data as a
JSON or
XML object. When different applications manipulate the same data in different formats,
synchronizing the applications amounts to synchronizing their respective data
formats. One way to synchronize data formats is to use \emph{lenses}, \emph{i.e.,} programs that represent pairs of functions for translating data from
format A to B (the \emph{get} direction) and back again (the \emph{put}
direction).  These lenses, when defined using domain-specific
languages like Boomerang~\cite{boomerang} or
BiGul~\cite{ko2016bigul}, typically come with a variety of guarantees,
the \emph{lens laws}, that help ensure put and get functions are effective
inverses of one another.
%comprise such synchronization functions, and the ``lens
%laws'' provide restrictions on how these synchronization functions interoperate%.
%Bidirectional languages like Boomerang~\cite{boomerang} and
%BiGul~\cite{DBLP:conf/pepm/KoZH16} use a single expression to encode all of the
%lens's synchronization functions at the same time.

@@

\noindent
Indeed, Boomerang's quotient lenses are structured this
way~\cite{quotientlenses}.
Of course, some of these elements are optional.  For instance, it is
not necessary to document internal and external formats.  However, we do
view it as good engineering practice to do so, as such documentation makes
lenses easier to read, understand, use and maintain.  In this case, there is
significant redundancy here:  External and internal formats
typically have some overlap; external formats and canonizers have related
structure; and the structure of bijective lens computations is largely driven
by the structure of the internal formats.  Hence, overall, a more compact
notation would simplify description and legibility of such computations.
In addition,
the lens and canonizer languages themselves can be somewhat fiddly and
difficult to
master.  For instance, to quotient out whitespace inside, and in between,
fields while also ignoring field order, one must nest canonizing combinators
within one another in a somewhat tricky way.

@@

One way to define the function used by squash is simply to write it
from scratch in OCaml.  However, our system will also synthesize
such functions automatically from pairs of regular languages and
examples.
%
\begin{lstlisting}
let l = synth comma_name $\Leftrightarrow$ space_name using {("Conway, J. H.", "J. H. Conway")}
let s_to_c = get l
\end{lstlisting}
%
The first line above synthesizes a \emph{bidirectional} lens between
\codefont{comma\_name} and \codefont{space\_name} using the listed example
transformation as a guide.  The second line simply
extracts the get direction transformation from the lens,
which is what we need for squash.

@@

\subsection{QRE Lenses}
QREs allow users to specify regular languages modulo equivalence relations,
a first step in writing many data transformations.  However, writing the
inner bijective lens itself still takes a fair amount of work.  Each such lens
is constructed from simple primitive combinators including:
%
\begin{itemize}
  \item \kw{del} $s$: delete the constant string $s$ in the get direction;
  produce the constant string $s$ in the put direction.
  \item \kw{ins} $s$: produce the string $s$ in the get direction;
  delete the string $s$ in the put direction.
  \item \kw{copy} $R$: copy the text matching the regular expression $R$ in both
  directions.
\end{itemize}
In addition, such primitives may be combined using the regular operators
Kleene star, alternation and concatenation, or functional composition of
lenses.  Figure~\ref{fig:example-lens} illustrates the use of these
combinators to transform \bibtex{} to EndNote (and vice versa).
A fair amount of trickiness in writing these transformations as users
must keep mental track of what they have parsed so far \emph{in both
directions}, carefully inserting and deleting the relevant syntax on both
sides.

@@

let author_lens : (lens in (",\n" . bib_author) $\Leftrightarrow$ ("\n" . end_author)) =
    del ",\nauthor = \""
        . ins "\n%A "
        . valid_name
        . (del " and "
        . ins "\n"
        . (ins "%A " . valid_name . del " and " . ins "\n")* 
        . ins "%A "
        . valid_name
        || "")

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.7\textwidth]{canonizers-outside}
\caption{Quotient Lens with ``Canonizers at the Edges''
\sam{We don't need a general choose function. With QREs the kernel language is
always contained in the whole language so we just use the inclusion as the
choose function.}\bcp{I can fix.}}
\label{fig:attheedges}
\end{figure}

@@

One combinator that we haven't yet introduced is sequential composition.
The $c \semicolon c'$ QRE successively applies
two equivalence relations to a regular language. For example, suppose that
we wish to express a name as $\kw{last\_name \cdot ``,\text{''} \cdot
\text{wsp+} \cdot first\_name}$, or as $\kw{first\_name\cdot \text{wsp+} \cdot
last\_name}$, with $\kw{first\_name\cdot \text{wsp+} \cdot last\_name}$ as the
canonical representation. We also wish to project away the extra whitespace in
between \kw{first\_name} and \kw{last\_name} once this identification has been
made. Then we can express this equivalence relation by a composition of a
\textit{squash} QRE and a QRE that projects away the extra whitespace in between
\kw{first\_name} and \kw{last\_name}:

@@

\noindent This rule says that the concatenation $c \cdot c'$ of QREs $c$ and
$c'$ is well formed only if the concatentions of $W(c)$ and $W(c')$, and
$K(c)$ and $K(c')$ are unambiguous.

@@

\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonizer(R) &=& id_{\mathcal{L}(R)} \\
\canonizer(\project{R}{s})(w) &=& s\\
\canonizer(\squash{c}{c'}{f})(w) &=&
\begin{cases}
\canonizer(c')(f(w)) & \text{if } w \in \mathcal{L}(W(c))\\
\canonizer(c')(w) & \text{otherwise}
\end{cases}\\
\canonizer(\normalize{R}{R'}{f}) &=& f\\
\canonizer(c' \; ; \; c) &=& \canonizer(c') \circ \canonizer(c)\\
\canonizer(c' \cdot c) &=& \canonizer(c') \cdot \canonizer(c)\\
\canonizer(c_1 \sep c_2)(w) &=&
\begin{cases}
\canonizer(c_1)(w) & \text{if } w \in \mathcal{L}(W(c_1))\\
\canonizer(c_2)(w) & \text{if } w \in \mathcal{L}(W(c_2))\\
\end{cases}\\
\canonizer(c^*) &=& \canonizer(c)^* \\
\end{array}
\]
\end{center}
\begin{align*}
&\canonizer(\perm{c_1, \ldots, c_n}{c})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})\\
&= (\canonizer(c_1)(r_1)) \cdot (\canonizer(c)(s_1)) \cdot \ldots \cdot
(\canonizer(c_n)(r_n))
\end{align*}
\caption{QRE canonizers}
\label{fig:canonizers}
\end{figure}

@@

\section{Conclusion}
\label{concl}
